---
title: "Exercises for ANOVA"
output: 
  html_document: 
    toc: true
    theme: readable

---


```{r, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, message=FALSE, error = TRUE)

```




---

> __Problem  statement:__


Compute an ANOVA for the `CO2` dataset. The effect of `Treatment` and `Type` on the carbon dioxide uptake rates should be examined. Give an interpretation for the result. Would you use a model with or without interaction?

---




Each of the two factor variables has two levels. In the first step we will compute mean values for each of the two levels

```{r}
library(tidyverse)
head(CO2)

```



Next we analyse if there might be an interaction between the two variables.

```{r int-co2}
CO2 |> 
ggplot(mapping = aes(x=Treatment,color=Type,group=Type,y=uptake)) +
stat_summary(fun.y = mean, geom = "point") +
stat_summary(fun.y = mean, geom = "line")

head(CO2)
bartlett.test(uptake ~ interaction(Treatment, Type),
data = CO2)
# significant p-value, there's interaction between treatment and type
summary(aov(uptake ~ interaction(Treatment, Type), data = CO2))
df <- aov(uptake ~ Treatment * Type, data = CO2)
library(ggfortify)
autoplot(df)

#method1
CO2$signal <- ifelse(CO2$conc > mean(CO2$conc),factor("high"),factor("low"))

bartlett.test(uptake ~ interaction(signal, conc),
data = CO2)
# there is significant interaction

#method2
CO2 <- CO2 |>
  mutate(conc_f = factor(
    ifelse()
  ))



```



There may be no interaction effect, but we check this by fitting the model with interaction.

```{r}
bartlett.test(uptake ~ interaction(Treatment, Type),
data = CO2)

```

The interaction effect is not significant at the 5% level.

```{r}
uptake_aov <- aov(uptake ~ Treatment * Type, data = CO2)
summary(uptake_aov)
```


---

> __Problem  statement:__


Think about the model assumptions. To check the assumptions, try using `autoplot()` on your ANOVA object.

---



```{r plot-co2-aov}
library(ggfortify)
autoplot(uptake_aov)

```







---

> __Problem  statement:__


In the second step, we want to check if the ambient carbon dioxide concentration `conc` affects `uptake` too. Check this with an ANOVA. The variable should be split at the median into two parts and coded as a `factor`, see `?cut`.



---

```{r}
#method1
CO2$signal <- ifelse(CO2$conc > mean(CO2$conc),factor("high"),factor("low"))

bartlett.test(uptake ~ interaction(signal, conc),
data = CO2)

#method2
CO2 <- CO2 |> 
  mutate(conc_f = factor(
    ifelse(conc <= median(conc),
           "low", "high"), levels = c("low","high")
    )
  )

uptake_aov2 <- aov(uptake ~ Type * conc_f + Treatment * conc_f, 
                   data = CO2)

summary(uptake_aov2)
CO2 |> 
  ggplot(mapping = aes(x = Treatment, color = conc_f, group = conc_f, y = uptake)) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun = mean, geom = "line")
```



Finally, let's fit the model **without interactions** and check the residuals.

```{r}
uptake_aov3 <- aov(uptake ~ Type + Treatment + conc_f, 
                   data = CO2)

summary(uptake_aov3)
```


```{r plot-co2-aov3}
autoplot(uptake_aov3)

```

